The seismic signature of upper-mantle plumes: application to the northern East-African Rift

### 2018-11-26 latex/manuscript.tex
Updates from ShareLaTeX
added a Vs figure and text updates

\author[1]{Chiara Chiviero}
\author[2]{John J.~Armitage}
\author[3]{Saskia Goes}
\author[4]{James O.~S.~Hammond}
%\author{[ngu]{Lo{\"i}c Fourel}}

@@

Only a few previous studies have tested the tomographic results against dynamic models to understand the nature and origin of mantle plumes, with very few using full resolution tests. \cite{boschi-etal-2007} and \cite{styles-etal-2011} did a comparison of global and regional plume models against global tomography to get an overview of how many plumes might exist. \cite{davaille-etal-2005} looked over a large region encompassing all of African and surrounding hotspots below the Atlantic and Indian Ocean, and qualitatively compared plume styles predicted from analogue models. Structures from dynamic plume models have been used by \cite{hwang-etal-2011} and \cite{maguire-etal-2016} to illustrate how wavefront healing may mask the travel-time signatures of plumes below 1000 km depth. \cite{ballmer-etal-2013} performed a regional study focusing on Hawaii and found that numerical models of a complex thermo-chemical plume are compatible with the tomographic images of the upper mantle below the islands.

@@

The Rayleigh B{\'e}nard convection develops uniformly. The Rayleigh Taylor destabilisation of a hot layer of material is however not uniform in time (models N7 to N10; see Fig.~\ref{fg:4}). Here we can see early and late stage plume-like structures within the same snap-shot. The spacing of these instabilities is of the order $\lambda = 0.5H$ to $\lambda = H$ depending on the aspect ratio of the model domain and the stage at which the plume forms (Fig.~\ref{fg:4}; see models N7 and N8). The first instability always forms at a corner of the model, and this subsequently leads to a destabilization of the hot layer that propagates outwards form the corner. The wavelength of the plume-like structures was found to be independent of the temperature of the hot layer, as this did not significantly affect the initial $Ra$ number. Furthermore, the contrast in temperature is high for these plumes when compared to the Newtonian Rayleigh B{\'e}nard convection (Fig.~\ref{fg:5}). The strong temperature contrast is likely important. When the temperature gradient is low the plume becomes less visible to seismic experiments \citep{styles-etal-2011}. Therefore given the spacing of the Rayleigh Taylor is similar to the Rayleigh B{'e}enard convection for the aspect ratio chosen, but the plumes have a stronger temperature contrast we will explore how they are transformed when viewed as seismic anomalies.

@@

Velocity anomalies are computed relative to a reference profile in a corner of the model (at $x=0$, $y=0$; e.g.~Fig.~\ref{fg:4}), an area with no upwelling or downwelling. The numerical models do not include the effect of melting. To include the basic effect of melt retention, we make a simple estimate of melt fractions and their signature: The dry solidus is approximated with a linear increase from $1100\,^{\circ}$C at 0\,km to $1700\,^{\circ}$C at 200\,km depth (based on the variation in the solidus from \citealp{herzberg-etal-2000}). Then we estimate a melt fraction from how much the actual (full) temperature exceeds the solidus, again using a linear relation where melt fraction is 0\,\% at the solidus temperature and reaches a maximum of 2\,\% for 300\,K excess temperature. We limit the range of melt fractions, assuming that melts are mobile and much larger melt fractions will escape rather than pool \citep[e.g.][]{faul-etal-1994}. To get the seismic anomalies due to melting, melt fractions are multiplied with the melt derivatives from \cite{hammond-2000}, which are -3.6 per percent melt volume for $V_{P}$ and -7.9 for $V_{S}$. This results in substantial local reductions in seismic wave speeds in the locations in the thermal structures where melt is expected.

@@

We focus on the Rayleigh Taylor instabilities given they have a strong temperature contrast (Fig. \ref{fg:5}), and have a wavelength that is similar to that interpreted form the tomographic observations. We will also focus on the models with an aspect ratio of 4x4x1, as the increased numerical resolution of these models (512x512x128) allowed us to model very large temperatures contrasts between the asthenosphere and the plumelet temperatures, 100 to 400 °C (models N8 to N10; Table \ref{tb:2})

@@

\begin{figure*}
\centering
\includegraphics[width=11cm]{../figures-working/fig07.png}
\caption{(a) P-wave velocity anomalies (\%) along a vertical cross-section through the input model N8 oriented such that the two plumes shown are positioned approximately under Afar and west of the MER. (b) Input S-wave velocity anomaly (\%). The location of the cross-sections (black line) is shown in Fig.~\ref{fg:1}b. The structure on the right represents the synthetic middle-stage (MS) plume, the structure on the left the late-stage (LS) plume discussed in the text. (c, d, g, h) Vertical cross-sections through the recovered undamped (damping $=0$) (c-d) P- (c) and S-wave (d) models and the moderately damped (damping $=35$) P- (g) and S-wave (h) models. e, i) Vertical cross-sections of the undamped (e) and damped (i) NEAR-P15 model. (f, j) Vertical cross-sections of the undamped (f) and damped (j) NEAR-S16 model. The spacing between the contours is 0.25\,\% for P-wave models and 0.50\,\% for S-wave models. White points indicate distance every $2^{\circ}$. Both the undamped models (c–d) and the damped recovered models (g–f) image most of the middle-stage plume, and only the head, but with relatively subdued amplitudes, of the late stage plume. The scale of the recovered structures is quite similar to that of imaged features.}
\label{fg:7}
\end{figure*}

We use the model N8 (see Table \ref{tb:2}) to illustrate the resolution of the synthetic plume structures below Afar and MER and compare the resolved models with the observed P- and S-wave tomography. We rotated the plume model in order to place a middle-stage plume with a broad head and a thick stem below Afar, a late-stage plume with a head spread at the base of the lithosphere, and a narrow tail below MER (Fig.~\ref{fg:7}a and b). Synthetic cross sections of the same orientation as the sections through models NEAR-P15 and NEAR-S16 in Fig.~\ref{fg:1} are shown in Fig.~\ref{fg:7}c, d, g, and h. The middle-stage plume located beneath Afar are generally well recovered, although details, such as a partially folded head and a tail influenced by phase-boundary topography are not identifiable. The anomalies from the late-stage upwelling below the MER are recovered less clearly, especially in the undamped case, where the plume head, which has spread along the base of the lithosphere, is not fully resolved (Fig.~\ref{fg:7}c, d). The recovery of the plume geometries can be enhanced if constraints on the shallow structure and damping towards it are added (Fig.~\ref{fg:7}g, h; see \citealp{civiero-etal-2016,civiero-etal-2015}). Resolution for the P- and S-wave models is visually very similar. Some of the differences between the recovery of the two plumes may be due to different resolution below Afar and MER as the data coverage is slightly higher in the first region than in the latter.

@@

@@

\begin{figure*}
\centering
\includegraphics[width=16cm]{../figures-working/fig09.png}
\caption{Vertical cross-sections through the undamped P-wave model from case A (model N9) focused in the region below Afar.  The orientation of the cross-sections (black line) is shown in the 200 km depth slices through the NEAR-P15 tomographic model  j. (a, d) synthetic and resolved images of the LS phase of plume evolution; (b, e) synthetic and resolved images of the MS phase of plume evolution; (c-f) synthetic and resolved images of the ES phase of plume evolution. The imaged EP (g-j) and LP (i-l) phases are similar although the synthetic structures (a, c) are different due to the lack of resolution at shallow upper-mantle depths. The MS phase (e) is well resolved because the head of the input model (b) is relatively large. g-i) Input and retrieved P-wave velocity LS (g), MS (h) and ES (i) anomaly envelopes (\%) along the green, blue and red profiles drawn in the cross-sections. Within the transition zone the retrieved and observed velocity anomalies overlap. (j) 200 km depth slice through the undamped NEAR-P15 model k) Vertical cross-section of the undamped NEAR-P15 model. The spacing between the contours is 0.25\,\%. White points indicate the distance every $2^{\circ}$.}
\label{fg:8}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=16cm]{../figures-working/fig10.png}
\caption{Vertical cross-sections through the undamped P-wave model from case A. (d-i) focused in the region below MER.  The orientation of the cross-sections (black line) is shown in the 200 km depth slices through the NEAR-P15 tomographic model  j.  (a, d) synthetic and resolved images of the LS phase of plume evolution; (b, e) synthetic and resolved images of the MS phase of plume evolution; (c-f) synthetic and resolved images of the ES phase of plume evolution. g-i) Input and retrieved P-wave velocity LS (g), MS (h) and ES (i) anomaly envelopes (\%) along the green, blue and red profiles drawn in the cross-sections. (j) 200 km depth slice through the undamped NEAR-P15 model k) Vertical cross-section of the undamped NEAR-P15 model. The spacing between the contours is 0.25\,\%. White points indicate the distance every $2^{\circ}$.}
\label{fg:9}
\end{figure*}

@@

Comparing forward models with observed tomography to find a good matching between synthetic and observed features is not straightforward. Tomographic resolution is spatially variable and highly non-linear, making it difficult to assess \citep{rawlinson-etal-2010}. However, from our set of resolution tests for each plume stage of evolution, we clearly recognize that models with a strong tail, like ES and LS plumes, do not match our observations well in character or amplitude profiles with depth (Fig.~\ref{fg:8} and \ref{fg:9}). LS plumes also do not fit the observed tomography as they do not produce sufficient transition-zone anomalies imaged below both Afar and MER (Fig. ??).  Plumes in a middle phase of evolution (MS) with a broad head and a short tail best explain the evolutionary stage of both the upper-mantle structures below Afar and MER (Fig.~\ref{fg:10}). A MS plume below Afar matches the amplitude of both the P- and S-wave low-velocity anomalies within the transition zone and above (Fig.~\ref{fg:10} q, r). Moreover, the similarity between the retrieved and observed tomographic shape of the plume is striking, especially for the S-waves (Fig.~\ref{fg:10}). The plume below MER could be in a slightly more advanced stage with a pronounced head spread at the base of the lithosphere and a quite thin ($\sim 100$\,km diameter) tail. Yet given the simplicity of the Cartessian input model, we would still suggest that the structure below the MER represents a middle stage plumelet.

@@

\bibliography{../../../../ref}

>>>>

\author[1]{Chiara Civiero}
\author[2]{John J.~Armitage}
\author[3]{Saskia Goes}
\author[4]{James O.~S.~Hammond}
%\author{[ngu]{Lo{\"i}c Fourel}}

@@

Only a few previous studies have tested the tomographic results against dynamic models to understand the nature and origin of mantle plumes, with very few using full resolution tests. \cite{boschi-etal-2007} and \cite{styles-etal-2011} did a comparison of global and regional plume models against global tomography to get an overview of how many plumes might exist. \cite{davaille-etal-2005} looked over a large region encompassing all of African and surrounding hotspots below the Atlantic and Indian Ocean, and qualitatively compared plume styles predicted from analogue models. Structures from dynamic plume models have been used by \cite{hwang-etal-2011} and \cite{maguire-etal-2016} to illustrate how wavefront healing may mask the travel-time signatures of plumes below 1000 km depth. MAGUIRE ET AL., (2018) To better understand plume resolution given the limitations of network design... \cite{ballmer-etal-2013} performed a regional study focusing on Hawaii and found that numerical models of a complex thermo-chemical plume are compatible with the tomographic images of the upper mantle below the islands.

@@

The Rayleigh B{\'e}nard convection develops uniformly. The Rayleigh Taylor destabilisation of a hot layer of material is however not uniform in time (models N7 to N10; see Fig.~\ref{fg:4}). Here we can see early and late stage plume-like structures within the same snap-shot. The spacing of these instabilities is of the order $\lambda = 0.5H$ to $\lambda = H$ depending on the aspect ratio of the model domain and the stage at which the plume forms (Fig.~\ref{fg:4}; see models N7 and N8). The first instability always forms at a corner of the model, and this subsequently leads to a destabilization of the hot layer that propagates outwards form the corner. The wavelength of the plume-like structures was found to be independent of the temperature of the hot layer, as this did not significantly affect the initial $Ra$ number. Furthermore, the contrast in temperature is high for these plumes when compared to the Newtonian Rayleigh B{\'e}nard convection (Fig.~\ref{fg:5}). The strong temperature contrast is likely important.

When calculated seismic structures are converted to synthetic tomography the inverted magnitude of $V_{S}$ or $V_{P}$ diminishes \citep[e.g.][]{goes-etal-2012,maguire-etal-2018}. Therefore a strong temperature contrast might be required to match the significantly low seismic velocities inverted for below Afar and the MER (Figure \ref{fg:1}). This would suggest the Rayleigh Taylor structures would more likely correspond to the observed tomography. The wavelength of the plumelet spacing is however in this case dependent on the model aspect ratio. For the mode N7 (3x3x1) the spacing between rising plumelets is close to 800\,km when scaled to the upper mantle. For models N8 through to N10 the spacing is closer to 1000\,km (Fig.~\ref{fg:4}). Given that the Rayleigh Taylor plumelets can match the observed spacing and have a stronger temperature contrast we will explore how they are transformed when viewed as seismic anomalies.

@@

\begin{figure*}
\centering
\includegraphics[width=11cm]{../figures-working/velocities.png}
\caption{Cross section of the S-wave seismic velocity anomaly relative to the horizontally averaged velocity for the Newtonian and non-Newtonian Rayleigh B{\'e}nard models N3 and N6 and the Rayleigh Taylor model N8.}
\label{fg:5b}
\end{figure*}

The plumelets can be seen in the seismic velocity anomalies to varying degrees depending on the model rheology and if they are due to Rayleigh B{\'e}nard convection (models N1 to N6; Fig.~\ref{fg:5b}a and b) or if  they form due to the destabilisation of a hot layer (models N7 to N10; Fig.~\ref{fg:5b}c). At shallow depths we also include the effect of melt. To include the basic effect of melt retention, we make a simple estimate of melt fractions and their signature: The dry solidus is approximated with a linear increase from $1100\,^{\circ}$C at 0\,km to $1700\,^{\circ}$C at 200\,km depth (based on the variation in the solidus from \citealp{herzberg-etal-2000}). Then we estimate a melt fraction from how much the actual (full) temperature exceeds the solidus, again using a linear relation where melt fraction is 0\,\% at the solidus temperature and reaches a maximum of 2\,\% for 300\,K excess temperature. We limit the range of melt fractions, assuming that melts are mobile and much larger melt fractions will escape rather than pool \citep[e.g.][]{faul-etal-1994}. To get the seismic anomalies due to melting, melt fractions are multiplied with the melt derivatives from \cite{hammond-2000}, which are -3.6 per percent melt volume for $V_{P}$ and -7.9 for $V_{S}$. This results in substantial local reductions in seismic wave speeds in the locations in the thermal structures where melt is expected.

@@

We focus on the Rayleigh Taylor instabilities given they have a strong temperature contrast (Fig. \ref{fg:5}), and have a wavelength that is similar to that interpreted form the tomographic observations. We will explore both models with an aspect ratio of 3x3x1 (model N7) and 4x4x1 (models N8 to N10; Table \ref{tb:2}). The increased numerical resolution of models N8 to N10 allowed us to model very large temperatures contrasts between the asthenosphere and the plumelet temperatures, 100 to $400\rm\,^{\circ}C$ (Table \ref{tb:2}).

@@

\begin{figure*}
\centering
\includegraphics[width=11cm]{../figures-working/fig07.png}
\caption{(a) P-wave velocity anomalies (\%) along a vertical cross-section through the input model N7 oriented such that the two plumes shown are positioned approximately under Afar and west of the MER. (b) Input S-wave velocity anomaly (\%). The location of the cross-sections (black line) is shown in Fig.~\ref{fg:1}b. The structure on the right represents the synthetic middle-stage (MS) plume, the structure on the left the late-stage (LS) plume discussed in the text. (c, d, g, h) Vertical cross-sections through the recovered undamped (damping $=0$) (c-d) P- (c) and S-wave (d) models and the moderately damped (damping $=35$) P- (g) and S-wave (h) models. e, i) Vertical cross-sections of the undamped (e) and damped (i) NEAR-P15 model. (f, j) Vertical cross-sections of the undamped (f) and damped (j) NEAR-S16 model. The spacing between the contours is 0.25\,\% for P-wave models and 0.50\,\% for S-wave models. White points indicate distance every $2^{\circ}$. Both the undamped models (c–d) and the damped recovered models (g–f) image most of the middle-stage plume, and only the head, but with relatively subdued amplitudes, of the late stage plume. The scale of the recovered structures is quite similar to that of imaged features.}
\label{fg:7}
\end{figure*}

We use the model N7 (see Table \ref{tb:2}) to illustrate the resolution of the synthetic plume structures below Afar and MER and compare the resolved models with the observed P- and S-wave tomography. We rotated the plume model in order to place a middle-stage plume with a broad head and a thick stem below Afar, a late-stage plume with a head spread at the base of the lithosphere, and a narrow tail below MER (Fig.~\ref{fg:7}a and b). Synthetic cross sections of the same orientation as the sections through models NEAR-P15 and NEAR-S16 in Fig.~\ref{fg:1} are shown in Fig.~\ref{fg:7}c, d, g, and h. The middle-stage plume located beneath Afar are generally well recovered, although details, such as a partially folded head and a tail influenced by phase-boundary topography are not identifiable. The anomalies from the late-stage upwelling below the MER are recovered less clearly, especially in the undamped case, where the plume head, which has spread along the base of the lithosphere, is not fully resolved (Fig.~\ref{fg:7}c, d). The recovery of the plume geometries can be enhanced if constraints on the shallow structure and damping towards it are added (Fig.~\ref{fg:7}g, h; see \citealp{civiero-etal-2016,civiero-etal-2015}). Resolution for the P- and S-wave models is visually very similar. Some of the differences between the recovery of the two plumes may be due to different resolution below Afar and MER as the data coverage is slightly higher in the first region than in the latter.

@@

The model N7 has a remarkable similarity to the observed seismic tomography (Fig.~\ref{fg:7}). When the model aspect ratio is 3x3x1, the plumelet spacing is of the order of 800\,km and this allows us to rotate two plumes into a position that matches the two low seismic anomalies found below Afar and the MER. Models N8 to N10 have a wider plumelet spacing, due to their larger aspect ration, 4x4x1. This larger aspect ratio was chosen due to the requirement to have a larger numerical resolution to solve for the destabilisation of a hot layer with increased temperature, while efficiently spreading the numerical model across compute nodes. Therefor to test if a hotter thermal anomaly is required to match the seismic anomalies, we will explore how individual plumelets potentially match the observed tomography.

@@

\begin{figure*}
\centering
\includegraphics[width=16cm]{../figures-working/fig09.png}
\caption{Vertical cross-sections through the undamped P-wave model from model N9 with a $200\rm\,^{\circ}C$ hot layer, focused in the region below Afar.  The orientation of the cross-sections (black line) is shown in the 200 km depth slices through the NEAR-P15 tomographic model  j. (a, d) synthetic and resolved images of the LS phase of plume evolution; (b, e) synthetic and resolved images of the MS phase of plume evolution; (c-f) synthetic and resolved images of the ES phase of plume evolution. The imaged EP (g-j) and LP (i-l) phases are similar although the synthetic structures (a, c) are different due to the lack of resolution at shallow upper-mantle depths. The MS phase (e) is well resolved because the head of the input model (b) is relatively large. g-i) Input and retrieved P-wave velocity LS (g), MS (h) and ES (i) anomaly envelopes (\%) along the green, blue and red profiles drawn in the cross-sections. Within the transition zone the retrieved and observed velocity anomalies overlap. (j) 200 km depth slice through the undamped NEAR-P15 model k) Vertical cross-section of the undamped NEAR-P15 model. The spacing between the contours is 0.25\,\%. White points indicate the distance every $2^{\circ}$.}
\label{fg:8}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=16cm]{../figures-working/fig10.png}
\caption{Vertical cross-sections through the undamped P-wave model from model N9 with a $200\rm\,^{\circ}C$ hot layer, focused in the region below MER.  The orientation of the cross-sections (black line) is shown in the 200 km depth slices through the NEAR-P15 tomographic model  j.  (a, d) synthetic and resolved images of the LS phase of plume evolution; (b, e) synthetic and resolved images of the MS phase of plume evolution; (c-f) synthetic and resolved images of the ES phase of plume evolution. g-i) Input and retrieved P-wave velocity LS (g), MS (h) and ES (i) anomaly envelopes (\%) along the green, blue and red profiles drawn in the cross-sections. (j) 200 km depth slice through the undamped NEAR-P15 model k) Vertical cross-section of the undamped NEAR-P15 model. The spacing between the contours is 0.25\,\%. White points indicate the distance every $2^{\circ}$.}
\label{fg:9}
\end{figure*}

@@

Comparing forward models with observed tomography to find a good matching between synthetic and observed features is not straightforward. Tomographic resolution is spatially variable and highly non-linear, making it difficult to assess \citep{rawlinson-etal-2010}. However, from our set of resolution tests for each plume stage of evolution, we clearly recognize that models with a strong tail, like ES and LS plumes, do not match our observations well in character or amplitude profiles with depth (Fig.~\ref{fg:8} and \ref{fg:9}).  Plumes in a middle phase of evolution (MS) with a broad head and a short tail best explain the evolutionary stage of both the upper-mantle structures below Afar and MER (Fig.~\ref{fg:10}). A MS plume below Afar matches the amplitude of both the P- and S-wave low-velocity anomalies within the transition zone and above (Fig.~\ref{fg:10} q, r). Moreover, the similarity between the retrieved and observed tomographic shape of the plume is striking, especially for the S-waves (Fig.~\ref{fg:10}). The plume below MER could be in a slightly more advanced stage with a pronounced head spread at the base of the lithosphere and a quite thin ($\sim 100$\,km diameter) tail. Yet given the simplicity of the Cartessian input model, we would still suggest that the structure below the MER represents a middle stage plumelet.

@@

\bibliography{ref}
